---
title: "How to Use Agentic AI for Responsible Archaeological Data Analysis in R"
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx"
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl"
abstract: |
  Archaeologists who use generative artificial intelligence (AI) for data analysis need procedures for inspecting code, evaluating methods, and checking sources. This how-to article describes three practices for responsible agentic AI use in R. First, develop enough familiarity with research tools and standards to judge AI-generated outputs. R supports analysis, Quarto combines code and text in executable documents, and Git records changes for inspection. Second, configure an AI agent in a local harness with explicit project instructions and permissions. Third, use skills, tools, and Model Context Protocol servers to research a question, critique a researcher-written plan, execute approved analyses, and verify results. We describe the inputs, actions, and checks for each practice, including precautions for restricted archaeological data. The workflow retains the researcher's responsibility for analytical decisions and produces code, documentation, and a report that others can inspect. We discuss setup costs, limitations of automated checking, and the distinction between reproducing a saved analysis and repeating an agent interaction.
keywords: |
  artificial intelligence; archaeology; reproducible research; R; data analysis
highlights: |
  - Learn R, Quarto, and Git to inspect analyses, regenerate reports, and review changes
  - Configure an agent harness with project instructions and explicit permissions
  - Use tested skills and tools to research, plan, execute, and verify analyses
---

# Introduction

Archaeologists who use generative artificial intelligence (AI) to assist with their work face a practical question: how to use these tools in a responsible and productive way? We define artificial intelligence as software that perform tasks associated with human intelligence, including communication, decision-making, and prediction [@GildeZuniga2024]. Generative AI uses patterns learned from training data to produce text, code, images, or other content [@Feuerriegel2024]. For the majority of archaeologists, the most commonly used generative AI tools are large language models (LLMs), such as ChatGPT and Claude, that we interact with via a web-based chat interface. While these models often generate useful content, outputs can also include hallucinations (content presented as factual without an adequate basis), invented citations [@Walters2023] and convincing-sounding incorrect explanations [@Leung2023; @Messeri2024]. Financial, environmental and energy costs, commercial dependence, and privacy also concern archaeologists using commercial LLM services [@Kansa2026]. AI currently has a jagged frontier, meaning that AI assistance improves performance on certain tasks but degrades quality on others of seemingly identical difficulty [@DellAcqua2026]. This makes it hard to predict when an AI will produce high-quality output, and necessitates fine-grained human oversight. In this guide we introduce a basic agentic AI workflow using open source software that is suitable for many kinds of archaeological work and has potential to address some of these issues. While traditional AI chatbots engage with humans in the form of turn-by-turn conversations on a website that require regular human interaction to complete a task, agents are a very different way to interact with AI. An AI agent can autonomously operate tools directly on our computer to complete multiple steps towards a specific goal without constant human interaction. This article describes how to use agentic AI to reduce the friction associated with many of the typical tasks of the archaeological research process that are done on a computer.

Agentic AI refers to systems in which a language model uses context (e.g. files on our computer) and tools (e.g. the R programming language) to plan and take actions (e.g. search the internet, edit our files), rather than only responding to a prompt [@Korinek2025]. An AI agent connects a language model to tools that can read files, execute code, or retrieve information. A harness is the software we install on our computer that manages those connections and the interaction with the researcher. A wide variety of agentic and harness systems have emerged recently, @tbl-taxonomies summarises these by increasing autonomy [@Haase2026; @Zimmer2026]. Our guide focuses on the low-autonomy end of that range: the agent works only within a single project and researcher directs each stage. We are not describing end-to-end fully automated AI pipelines that generate ideas, run experiments, write manuscripts, and review the results with limited human intervention [@Lu2026]. Instead, we recommend a configured assistant for particular routine research tasks, whose file access, code execution, and outputs remain subject to researcher evaluation and approval.

```{r}
#| label: tbl-taxonomies
#| tbl-cap: "Levels of agentic AI in two recent taxonomies."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-taxonomies.csv")), col.names = c("Level", "Haase et al. (2026)", "Zimmer et al. (2026)"))
```

In this researcher-directed workflow, we follow the "human at the helm" approach to describe responsibility for setting goals, approving methods, interpreting findings, and deciding what to publish [@Monakhov2026; @Jonker2024]. This can be contrasted with human-in-the-loop approaches where human intervention is mostly as a validator rather than decision-maker [@Zanzotto2019; @Taylor2023]. @tbl-helm-vs-loop compares the two approaches. With the human at the helm, control requires observable actions: the researcher reviews changes, checks evidence, and can stop or redirect the work. In our view, human engagement and control remain essential even as technical capabilities advance [@Spillias2025].

A high level of researcher control also necessary to honor our responsibilities toward the communities involved in our work and affected by it. AI-generated perspectives cannot replace community participation [@Spillias2025]. Where archaeological work requires community consultation or consent, our approval of an agent's actions does not replace those obligations. Recent guidance on archaeological AI ethics and heritage management provides additional reasons to make those responsibilities explicit [@Tenzer2024; @Shaikhon2025].

```{r}
#| label: tbl-helm-vs-loop
#| tbl-cap: "Human in the loop versus human at the helm across eleven dimensions."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-helm-vs-loop.csv")))
```

We translate these responsibilities into three practices for using agentic AI in archaeological work. First, develop enough familiarity with research tools and standards to evaluate the quality of AI-generated outputs. Second, configure an AI agent in a local harness to work with project files under researcher-defined permissions. Third, use instructions, skills, and Model Context Protocol (MCP) servers to research, plan, execute, and verify data analyses. These practices will be most useful to archaeologists who already analyze their own data in R, whether in academia, government, or private practice, and are looking to reduce some of the friction and tedium in their current workflow. 

# Step 1: Learn the Research Tools and Standards {#sec-tools}

We recommend learning to import, analyse and visualise data in R before involving an agent in an analysis; computational literacy must be a core methodological competence for data-intensive reseachers. You need to be able inspect agent-generated code and explain its relationship to the archaeological question. Learning these fundamentals provides a basis for evaluating the quality and relevance of AI suggestions [@Dyck2026; @Bridgeford2026]. Our view reflects the current consensus in many fields that learning coding fundamentals remains essential despite the expanding availability of AI assistance [@Campbell2024; @Dyck2026; @Abramson2026]. Cooper and colleagues also reaffirm teaching coding so researchers can interpret generated code and think critically about it, while noting variable output quality in mathematics, logic, and reproducibility [@Cooper2024; @Nejjar2025]. 

We recommend choosing a small set of reputable tools and learning how they work together, for example the tidyverse for data preparation and ggplot2 for visualization [@Wickham2019]. The here package helps construct project-relative file paths, and the renv records R package dependencies. The Quarto document format is ideal for writing your report, this combines prose, executable code, and results in a document written as a plain-text `.qmd` file [@QuartoDocs2026]. Quarto is an extension of the Markdown document format, which LLMs read and write very efficiently and accurately, much more so than Word documents and PDFs. This article was written as a Quarto document, the original file may be inspected in our research compendium at https://doi.org/00.0000/0000. We recommend using converters to turn DOCX and PDF files into markdown before agents read them [@Anthropic2025Skills]. You should create documents in Markdown format that you want your AI agent to read and edit. These R packages and document types also have the advantage of supporting the organization of transparent, reproducible and reusable analyses [@Marwick2017].

Git is another free and open source software that specializes in collaboration and recording versions of project files (e.g. text, code, data) so you can inspect changes and retrieve earlier work [@Wilson2014]. Version control with Git supports reproducibility and transparency in science [@Ram2013; @Blischak2016]. Git records commits, a commit is a saved snapshot of tracked files, accompanied by a message describing the changes; a diff shows differences between versions. Commits should be kept small, guidance for scientific computing recommends reviewing changes in small increments [@Wilson2014]. The Git history for this article can be inspected at https://github.com/benmarwick/agenticai. Agents are fluent in Git and very efficient at using it to track changes over multiple files in a project. The toolkit we have sketched out here, centered on R, Markdown, and Git, is especially well-suited for working with AI agents because it gives the agent and the researcher a shared, inspectable medium using free and open source tools. Agents read Markdown instruction files, run the R code embedded in our Quarto documents, and their edits are traceable as Git commits. This is a simple and robust system for linking reported results back to the executable analyses that generated them. Every machine-generated change is recorded transparently in plain text where you can review before accepting or rejecting it.

# Step 2: Configure an Agent Harness for Your Project {#sec-harness}

When we use the tools described in @sec-tools, it is relatively simple to configure an AI agent so that it can efficiently see all the files in our project and work on them. The key tool here is a local harness, this is software you install on your computer that turns a LLM into an agent. The LLM by itself can only generate text. The harness connects that model to tools, such as reading and editing project files, running shell and R commands, calling MCP servers, and managing the exchange between the model, the tools, and you. The harness enforces permissions (decides which actions run automatically and which require your approval, according to your configuration of the harness, e.g. which files they can and cannot edit), and manages context (assembles what the model sees each turn in the conversation, e.g project instructions such as AGENTS.md, conversation history, file contents, and tool outputs). It also runs the loop (sends your prompt to the model, sends the model's action requests to the tools, feeds results back to the model, and repeats until the task finishes or you stop it), and presents the interface (the space where you type prompts, watch the agent work, and read its output). At the time of writing widely used harnesses include Claude Code, GitHub Copilot, OpenAI Codex and OpenCode. We use and recommend OpenCode (running in the terminal pane of RStudio) because it is free and open source and not tied to any model vendor.

Store preferences and working habits you reuse across projects in a global AGENTS.md file. In OpenCode this file is located in a path similar to .config/opencode/AGENTS.md. Keep project-specific details in a second AGENTS.md file located at the top level of each project folder. This project-specific preferences file is the place to record contextual information specific to the project you are currently working on, and saves you having to manually send those details to the LLM in each session. This could include information about archaeological context, about prior decisions you have made about the analysis plan, about your target audience for your report or manuscript, etc. We provide examples of our AGENTS.md files in our research compendium.

Harnesses such as OpenCode often provide multiple working modes, 'plan' and 'build' are two widely used modes. Plan directs the model to inspect files and propose changes without editing any project files. Build allows file editing and running code according to your permission settings. We recommend using Plan mode when exploring options with the model and drafting your analysis plans. This prevents the model from making changes to your files before your plan is finalized and that may be complicated to undo. We recommend multiple rounds of planning with a model to explore options for you to choose from before you decide on a course of action. First-pass AI suggestions can be unreliable because models can produce authoritative-sounding but wrong proposals and illusions of understanding [@Messeri2024]. A model's conclusions can shift with model and prompt choices [@Bertran2026], so each round needs evaluation against data, literature, and your expertise. The specificity of a plan compounds across rounds, with each discussion round letting you add variables, design details, and constraints missing from previous rounds [@Brown2026b]. Multiple rounds of planning also gives you a chance to steer and challenge the model, request justification for decisions it proposes, and invite the model to criticize your ideas, in anticipation of human reviewers [@Korinek2023]. This helps to surface forking paths (alternative analytical decisions that could lead to different conclusions) early and map out the analysis space available to your project. We recommend saving your analysis plan to a markdown file at the end of a planning session (switch to build mode to allow the model to create a new file). This is because models often perform increasingly poorly across long meandering threads and rarely recover after a wrong turn [@Laban2025]. Then when you resume your work, you should start a new conversation by directing the model to read your plan saved in the markdown file, rather than rely on its memory. Having your plan in a markdown file also allows you to easily directly edit it independent of the AI, and manually record the status of each step of the plan (e.g. 'done', 'in progress'), so you remain at the helm of your project.

Planning an analysis with a LLM can be effective when using a state-of-the-art, reasoning-enabled large frontier model because these are often better at the long-horizon, open-ended, branching tasks involved in drafting a plan [@Zhang2025; @Lu2026]. Because larger models can be slower and more expensive, we recommend switching to a smaller model to implement the individual steps of the plan. When your plan is set, switch your harness to build mode and direct the model to execute one step of the plan. The more precisely specified the plan step is, the better chance you have of the smaller model successfully executing it [@Brown2026b]. As the model completes a step, verify completion by inspecting the Git diff, check the output against your plan, and for more complex steps, interrogate the model to have it justify its actions. Agents sometimes report finished work they never completed [@Brown2026a; @Nejjar2025], so it is essential check the files and outputs yourself.

While the harness is installed and runs locally on our computer, they typically use LLMs hosted remotely at data centers. This connection requires obtaining an API (Application Programming Interface) key from a model provider, and adding this key to our harness configuration file. Once we have this connection, our file contents and tool outputs may enter the context sent to a remote model, and potentially even be used to train future LLMs. We have been using the free tiers of models available from OpenRouter and OpenCode. However, for many archaeological projects using models hosted by commercial providers is unacceptable, for example when working with restricted site coordinates, confidential client information, and culturally sensitive data. One solution to this problem is to select LLM providers that offer an option for Zero Data Retention (ZDR), a contractual policy promise where the provider processes your request and deletes the data immediately after. Another option is to keep sensitive files in a directory entirely outside the project folder so the agent cannot open them by accident [@Brown2026book]. Simulated data with the same structure can replace restricted data during analysis development. Finally, perhaps the most secure solution is a locally hosted LLM so that no data leave your computer [@Abramson2026; @Wiest2024]. Widely used free and opensource tools for managing local LLMs include Ollama, llama.cpp, vLLM, and AnythingLLM, and currently popular models for local use include Llama (Meta), Gemma (Google) and Qwen (Alibaba). Local LLMs can suit simpler tasks when run on a typical computer, such as execution of well-specified steps of a plan, and writing code, but getting performance similar to leading hosted models requires expensive high performance hardware [@Brown2026local]. A hybrid approach uses a local LLM to inspect the raw data to prepare a synthetic version of that data that has exactly the same structure as the real dataset (same columns, same variable types, same format), has the same statistical properties as the real dataset, but contains no real locations, measurements, and no identifying information. This synthetic proxy data can then sent to the more capable remote model to write code that can be run locally on your data on your computer [@Schweinberger2026privacy]. 

# Step 3: Use Tools, Skills, and Model Context Protocols to Research, Plan, and Analyze {#sec-analysis}

With the harness configured, working with an AI to generate plans and taking actions is more efficient by using tools, skills and Model Context Protocols (MCPs) to do common research tasks such as search for and verify references, write and run code, organise and manage files (@Anthropic2024), and simulate peer review. A tool performs a single computer action and returns a result, such as reading, editing and renaming a file, or starting R and running R code. A skill is a markdown file of metadata and instructions and supporting resources for a recurring task [@Xu2026]. The agent reads the full skill file when a task contains words that matches the metadata, and the skill file contents is sent to the model as part of the prompt. For example, we have a skill file for the tidyverse idiom of the R language. This specifies conventions that we want the model to follow when it writes R code, e.g. use the native pipe |> over %>%. We also provide good and bad code examples. When we ask for R code, the harness loads that file into the context and sends it to the model along with the rest of the prompt, and the output follows those conventions. MCP is an open source standard for connecting an agent to external data sources and services; they can also provide specialist tools [@Anthropic2024; @Hou2025]. A skill describes a procedure, while tools perform its actions and an MCP server supplies connections. @tbl-stack lists the five MCP servers and six skills named in this guide plus the three instructions files that govern them.

```{r}
#| label: tbl-stack
#| tbl-cap: "MCP servers and skills installed in the author's OpenCode configuration."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-stack.csv")))
```

As as example of how these additions work, consider what happens if we give our agent the command "deeply research the debate about the Châtelperronian". This will load the deep-research *skill*, whose trigger keywords include that phrase. This specific skill contains instructions for a specific set of actions: scoping the authorship question, systematic search, source verification, synthesis with a devil's-advocate pass, and a written report. The harness dispatches subagents for search, verification, and synthesis; retrieval *tools* fetch candidate papers while verification tools check each citation against publication databases through a reference-verification server. The literature search and verification tools are provided by the refcheck MCP server; they query Crossref, Semantic Scholar, and other online services. The results of the literature search can then be written into the project-level bibtex file or saved in your Zotero library using the zotero-cli *skill* which drives the shell *tools* to connect with your Zotero database. Skills guides the agent but cannot force compliance, so you need to critically review the results before incorporating them into your work. We recommend multiple iterations of literature search prompts, adjusting the prompt each time to the steer the model to focus on different areas of the literature [@Moorthy2026]. Multiple rounds of verification prompts are also important to ensure the bibliography is free of hallucinated details.

Aided by tools, skills and MCP servers, AI agents can be very efficient at accurately extracting data from the full text of publications. One LLM extracted ecological data from literature 50 times faster than a human reviewer, with accuracy above 90% on discrete and categorical data [@Gougherty2024]. @Hariri2025 describes structured knowledge extraction from archaeological texts, and @Syme2026 examines validation of LLM-generated units for cross-cultural anthropological research. Human validation is necessary: @Gilardi2023 found {LLM} labels matched or exceeded crowd workers on tested tasks, @Ziems2024 recorded uneven validity across social concepts, and @Gougherty2024 observed that AI performed poorly on certain quantitative data, so quantitative fields need extra quality checks. GPT-4 extracted species at F1 99.8% and locations at 95.3% from pest-control abstracts and distinguished ecological roles, though a small number of confabulated details appeared [@Scheepens2024]. Context-rich prompting improves data extraction results [@Moorthy2026], though no single model dominates all extraction tasks, so you should test several model-prompt pairings and keep validation workflows that build trust in outputs. Before using machine-coded variables, you should define them in detail, compare a AI-collected sample against human coding, and document errors. Save source locations with extracted values so another researcher can inspect the evidence.

Although writing code from plain English prompts is perhaps the single greatest strength of most LLMs [@Zheng2024], generating high-quality, real-world code for research still requires your understanding of context, debugging skills, and integration with the overall goals of your research project. When using an AI agent to plan and write code for data analysis, we recommend @Brown2026b's guidance: 

- (1) write a prompt that is detailed and specific about the study's context, research question, types of variables and data collection methods. The archaeological expertise you bring to writing this prompt can make a big difference to the quality and consistency of responses to this prompt [@ZamfirescuPereira2023]. Rich context minimizes ambiguity and helps to align the model with your goals.
- (2) include with the prompt the dataset, a narrative summary of the data or a synthetic version of the data generated by a local LLM. This is another form of context and will improve the quality of the model's output [@Jansen2025]. 
- (3) include high-quality trustworthy reference materials, such as DOIs to exemplary studies similar to what you want to do, or technical methodological literature that describes relevant methods in detail. Use the agent to help you deeply search for this literature.
- (4) include instructions for the model evaluate its code for correctness and completeness before stopping. This sets the agent to work in a loop: generate code -> run it -> test it -> fix any errors -> run it again -> test it, etc. until your evaluation criteria have been satisfied. Specifying evaluation criteria and instructing the model to debug itself improves the accuracy of output [@Chen2023]



Tools that execute R can expose errors for inspection, but automated correction still needs review. @Brown2026a tested agentic workflows for ecological and fisheries modelling and found variable accuracy and failures to follow instructions. When code fails, check whether the proposed correction preserves the approved method. Do not accept a different analysis simply because it runs. Agents also change methods without flagging it: one agent swapped a requested bootstrap interval for a t-distribution interval, so read the code produced, not the summary [@Brown2026a].

# Discussion

The three practices assign distinct responsibilities to the researcher. Familiarity with R, Quarto, and Git supports efficient interaction with AI agents, supporting your inspection of calculations, reports, and changes. Harness configuration is where you define the AI agent's permissions and work habits. Use of tools, skills, and MCP servers greatly expand the capability of the model, and improve the accuracy and usefulness of its work. This sequence applies existing scientific-computing practices to agent-assisted archaeological work [@Wilson2014; @Marwick2017; @Bridgeford2026].

Each practice has costs. Learning R and version control requires time, while configuring a harness introduces provider choices and access decisions. Skills and servers add dependencies that require testing and maintenance. @Kansa2026 discusses commercial dependence, and @JohnsonEberl2026 examines environmental costs of AI in archaeology. @Luccioni2024 measured inference energy across models and tasks, so we batch queries and log usage. Treat heavy token use as a reason to inspect output: in one study the costliest runs were agents stuck in unproductive loops, not the most accurate ones [@Brown2026a]. Mapping latency, token cost, and F1 score guides model choice under budget constraints [@Moorthy2026]. We recommend recording available cost and usage information, limiting unnecessary repeated calls, and considering whether an existing R script already completes the task.

These instructions do not establish that agent assistance improves every analysis. The illustrative tasks are teaching examples, and we have not measured their effects on accuracy or completion time. Reproducing saved R code differs from repeating an agent conversation: a model may generate different code on another attempt. Provider-side model updates add a second source of non-repeatability: the same prompt can behave differently months later because the provider changed the model [@Abdurahman2025; @Abramson2026]. Preserve the inspected code and document the interaction, rather than promising identical agent behavior. Incentives sharpen a further risk: when publication metrics define success, researchers may use LLMs as a shortcut to game output rather than support innovation, which reduces quality and stifles development [@Johnson2024].

# Conclusion

 automate micro-tasks for productivity gains; comparative advantage is AI generates, human evaluates; do not over- or under-estimate LLMs because authoritative style does not equal truth; prompt with context, iterate, 

We recommend three practices for responsible agentic AI use in archaeological data analysis: learn the research tools and standards, configure a harness with explicit permissions, and use tested skills and tools to research, plan, and analyze. Quarto makes the relationship between code and report inspectable, while Git records reviewed changes. Planning and verification remain part of the researcher's work throughout.

The practical result should be a documented analysis whose inputs, decisions, code, and outputs another researcher can examine. The agent may assist with each task, but the researcher must explain the methods, assess the evidence, and authorize release. These responsibilities define when the work is ready to share.

# Acknowledgments

No permits were required for the preparation of this manuscript. The author used OpenCode (https://opencode.ai/), an open-source AI harness, and OpenRouter (https://openrouter.ai/) a commercial model provider. AI models used included Nemotron 3.5 Ultra (NVIDIA), Qwen3.8, Muse Spark 1.3 (Meta), GLM-5.3 Flash, and MiMo-V2.5 (Xiaomi). AI was used to assist assist with searching for literature, planning, drafting and editing the manuscript, and reference checking. All AI-generated suggestions were reviewed and revised by the author.

# Funding Statement

This research received no specific grant funding from any funding agency, commercial or not-for-profit sectors.

# Data Availability Statement

No original data were used.

# Competing Interests

The author declares none.



# References

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
