---
title: "A Practical Guide to Responsible Agentic AI for Archaeological Data Analysis in R"
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx"
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/society-for-american-archaeology.csl"
abstract: |
  Archaeologists who use generative artificial intelligence (AI) for data analysis need procedures for inspecting code, evaluating methods, and checking sources. This how-to article describes three practices for responsible agentic AI use in R. First, develop enough familiarity with research tools and standards to judge AI-generated outputs. R supports analysis, Quarto combines code and text in executable documents, and Git records changes for inspection. Second, configure an AI agent in a local harness with explicit project instructions and permissions. Third, use skills, tools, and Model Context Protocol servers to research a question, critique a researcher-written plan, execute approved analyses, and verify results. We describe the inputs, actions, and checks for each practice, including precautions for restricted archaeological data. The workflow retains the researcher's responsibility for analytical decisions and produces code, documentation, and a report that others can inspect. We discuss setup costs, limitations of automated checking, and the distinction between reproducing a saved analysis and repeating an agent interaction.
keywords: |
  artificial intelligence; archaeology; reproducible research; R; data analysis
highlights: |
  - Learn R, Quarto, and Git to inspect analyses, regenerate reports, and review changes
  - Configure an agent harness with project instructions and explicit permissions
  - Use tested skills and tools to research, plan, execute, and verify analyses
---

# Introduction

Archaeologists who use generative artificial intelligence (AI) to assist with their work face a practical question: how to use these tools in a responsible and productive way? We define artificial intelligence as software that perform tasks associated with human intelligence, including communication, decision-making, and prediction [@GildeZuniga2024]. Generative AI uses patterns learned from training data to produce text, code, images, or other content [@Feuerriegel2024]. For the majority of archaeologists, the most commonly used generative AI tools are large language models (LLMs), such as ChatGPT and Claude. While these models often generate useful content, outputs can also include hallucinations (content presented as factual without an adequate basis), invented citations [@Walters2023] and convincing but incorrect explanations [@Leung2023; @Messeri2024]. Financial, environmental and energy costs, commercial dependence, and privacy are also concerns for archaeological use of mainstream commercial LLM services [@Kansa2026]. The jagged frontier of AI means AI assistance improves performance on certain tasks but degrades quality on others of seemingly identical difficulty, making it hard to predict when an AI will produce high-quality output [@DellAcqua2026]. In this guide we introduce a basic agentic AI workflow suitable for archaeological work that has potential to address some of these issues.

Agentic AI refers to systems in which a language model uses rich context and specialised tools to plan and take actions, rather than only responding to a prompt. An AI agent connects a language model to tools that can read files, execute code, or retrieve information. A harness is the software we install on our computer that manages those connections and the interaction with the researcher. A wide variety of agentic and harness systems have emerged recently, @tbl-taxonomies summarises these by increasing autonomy [@Haase2026; @Zimmer2026]. This guide focuses on the low-autonomy end of that range: the researcher directs each stage, and the agent works only within a single project. We are not describing end-to-end fully automated AI pipelines that generate ideas, run experiments, write manuscripts, and review the results with limited human intervention [@Lu2026]. Instead, we recommend a configured assistant for particular tasks, whose file access, code execution, and outputs remain subject to researcher evaluation and approval.

```{r}
#| label: tbl-taxonomies
#| tbl-cap: "Levels of agentic AI in two recent taxonomies."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-taxonomies.csv")), col.names = c("Level", "Haase et al. (2026)", "Zimmer et al. (2026)"))
```

In this researcher-directed workflow, we use "human at the helm" to describe responsibility for setting goals, approving methods, interpreting findings, and deciding what to publish [@Monakhov2026; @Jonker2024]. Human-in-the-loop approaches also include human feedback and intervention [@Zanzotto2019; @Taylor2023]; neither label alone establishes meaningful control. @tbl-helm-vs-loop compares the two framings dimension by dimension. In this workflow, control requires observable actions: the researcher reviews changes, checks evidence, and can stop or redirect the work. In a perspective on future automated ecosystem modeling, @Spillias2025 argues that human engagement and control remain essential even as technical capabilities advance.

Researcher control also requires attention to responsibilities toward affected communities. @Spillias2025 cautions that AI-generated perspectives cannot replace participation by affected communities. Where archaeological work requires community consultation or consent, researcher approval of an agent's actions does not replace those obligations. Guidance on archaeological AI ethics and heritage management provides additional reasons to make those responsibilities explicit [@Tenzer2024; @Shaikhon2025].

```{r}
#| label: tbl-helm-vs-loop
#| tbl-cap: "Human in the loop versus human at the helm across eleven dimensions."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-helm-vs-loop.csv")))
```

We translate these responsibilities into three practices for using agentic AI in archaeological work. First, develop enough familiarity with research tools and standards to judge AI-generated outputs. Second, configure an AI agent in a local harness to work with project files under researcher-defined permissions. Third, use skills, tools, and Model Context Protocol (MCP) servers to research, plan, execute, and verify data analyses. These practices will be most useful to archaeologists who currently analyze their own data in R, whether in academia, government, or private practice, and are looking to reduce some of the friction and tedium in their current workflow. 

# Step 1: Learn the Research Tools and Standards {#sec-tools}

We recommend learning to import data, calculate summaries, and draw a figure in R before asking an agent to modify an analysis. Researchers also need to inspect the code and explain its relationship to the archaeological question. Learning these fundamentals provides a basis for evaluating AI suggestions [@Dyck2026; @Bridgeford2026]. @Vaithilingam2022 found that code-assistant users still struggled to verify generated code, so we require expert review of every AI-generated change before accepting it.

Choose a small set of tools and learn how they work together. We use the tidyverse for data preparation and ggplot2 for visualization [@Wickham2019]. The here package helps construct project-relative file paths, and renv records R package dependencies. These tools support the organization of reproducible analyses, although they do not guarantee that code will run in every environment.

Quarto combines prose, executable code, and results in a document written as a plain-text `.qmd` file [@QuartoDocs2026]. Quarto works the same way as R Markdown, combines its bookdown and distill extensions, and runs the Knitr and Jupyter engines [@QuartoFAQ2026]. Practice reading a data file, drawing a figure, and rendering that document before delegating changes to an agent. Check that changing the code changes the corresponding result in the rendered report. Keep the `.qmd` source as the document you edit; regenerate Word or HTML outputs from it. Markdown is the most efficient format for LLMs to read and write: skills ship as markdown instruction files, and converters turn PDFs into markdown before agents read them [@Anthropic2025Skills].

Git records versions of project files so researchers can inspect changes and retrieve earlier work [@Wilson2014]. Version control with Git supports reproducibility and transparency in science [@Ram2013], and primers introduce the Git and GitHub workflow to new users [@Blischak2016]. A commit is a saved snapshot of tracked files, accompanied by a message describing the changes; a diff shows differences between versions. Before inviting an agent to edit a project, practice inspecting a diff and making a commit of reviewed changes. Save a known working version of the Quarto document and its supporting code, then inspect each agent-generated change. Keep credentials and restricted data out of any shared repository.

These skills require practice. @Campbell2024 and @Dyck2026 argue for learning coding fundamentals despite the availability of AI assistance. Cooper and colleagues also reaffirm teaching coding so researchers can interpret generated code and think critically, while noting variable output quality in mathematics, logic, and reproducibility [@Cooper2024]. Sociologists now describe computational literacy as a core methodological competence rather than an optional extra [@Abramson2026]. @MorganWright2018 examines the relationship between drawing and interpretation in archaeological field recording. We use this as an analogy for attending to the work a tool performs, rather than evidence that manual coding always improves understanding.

Consider a task that counts artifacts by site. Before requesting code, specify that each row represents one artifact, explain missing site labels, and decide how the plot should display counts. A short base R solution might use `table()` and `barplot()`. A tidyverse solution might use `count()` and `geom_col()`. The package choice alone cannot establish correctness: the researcher must check what each operation counts and excludes.

To practice, use a small, non-sensitive data file whose counts you can check by hand. Write or adapt code from a tutorial, explain how it handles missing labels, and compare the plotted totals with the included records. Save these checks for evaluating later agent-generated changes. Guidance for scientific computing recommends testing programs against known results and reviewing changes in small increments [@Wilson2014].

Before proceeding, render the Quarto document from a new R session and inspect its figure. Commit the reviewed source files and record the expected counts in the project documentation. The output of this practice is an analysis the researcher can explain, a reproducible document, and a Git history against which later changes can be compared.

# Step 2: Configure an Agent Harness for Your Project {#sec-harness}

Use the project prepared in @sec-tools to configure the agent's working environment. We recommend a local harness when the task requires editing and running project code. The relevant distinction is the configured access to files and tools: a text-only exchange requires manual transfer, whereas a tool-enabled interface may act on the project. @tbl-advantages lists capabilities to check before choosing software.

```{r}
#| label: tbl-advantages
#| tbl-cap: "Capabilities to check when selecting and configuring an agent harness. Availability depends on the software, model, and permissions."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-advantages.csv")))
```

## Choose a Harness and Model

The harness and model have different roles. The harness manages files and tool calls; the model generates responses and requests actions. OpenCode provides configurable permissions for reading, editing, and executing commands [@OpenCodePermissions2026]. Aider offers a different workflow in which users select files for editing and inspect the resulting changes [@AiderUsage2026]. Compare the software's documentation against the tasks in your project rather than assuming that every harness supports the same capabilities.

For either route, install the software from its documented source, select a supported model, and follow the provider's authentication instructions. An application programming interface (API) key is a credential that authorizes access to a service; keep it outside version-controlled files. OpenCode users should configure permissions before requesting changes. Aider users should select only the files needed for the task and review its automatic-commit settings. Record the harness version, model identifier, provider, and date, and set a spending limit where available.

A local harness does not ensure that data remain local. OpenCode's documentation distinguishes local processing from API calls to the model provider [@OpenCodeEnterprise2026]. File contents and tool outputs may enter the context sent to a remote model. Before connecting, establish which materials may leave the institution and which services are approved. Exclude restricted site coordinates, confidential client information, and community-governed knowledge unless the relevant permissions explicitly allow their use. Keep sensitive files in a directory entirely outside the project folder so the agent cannot open them by accident [@Brown2026book]. Prefer a provider with a no-data-retention option where available, and never paste analysis outputs back into prompts, which leaks the same information as the raw data [@Brown2026book]. Simulated data with the same structure can replace restricted data during code development [@Brown2026book]. A locally hosted model suits simpler tasks on capable hardware but trails leading hosted models on harder ones [@Brown2026book]. One sociological team shared no data with commercial systems and extended codes with local models in under one-third the manual time, with secondary human review ensuring accuracy [@Abramson2026].

## Set Project Instructions and Permissions

Write a short project instructions file that states the conventions established in @sec-tools. In OpenCode, `AGENTS.md` supplies project instructions [@OpenCodeRules2026]. Include the language, document format, file locations, and checks expected before reporting completion. For example, instruct the agent to use R, edit the Quarto source, preserve raw data, explain proposed dependency changes, and request approval before committing or publishing files.

Instructions need technical controls as well as prose. OpenCode can allow, deny, or request approval for specified actions [@OpenCodePermissions2026]. Start with restricted access and require approval for edits and commands until you understand their effects. Treat `AGENTS.md` as guidance the agent may fail to follow, and check its actions against the configured permissions. Maintain a separate backup: Git can recover committed versions, but cannot recover every unsaved or untracked file.

## Test the Connection

Begin with a read-only task using non-sensitive documentation. Ask the agent to summarize the research question, name the data files, and identify the variables it would need. Compare the response with the actual files. Check the permission settings directly; an agent's claim that it lacks access is insufficient evidence of a restriction.

Next, use a disposable copy of the project to test a small authorized edit. Ask the agent to change a figure label, inspect the diff yourself, and render the document. Confirm that the intended label changed and that the calculation remained intact. The output is a configured agent, written conventions, and a tested procedure for reviewing its changes. If the agent's answers go wrong, restart with a clearer prompt in a fresh conversation rather than patching the same thread: an early wrong answer keeps shaping later replies [@Brown2026book]. The penalty is large: across 200,000 simulated conversations, models performed 39% worse in multi-turn than single-turn settings and rarely recovered after a wrong turn [@Laban2025]. Delete superseded scripts instead of keeping several versions of the same file: stale code in the project can keep influencing suggestions [@Brown2026book].

# Step 3: Use Skills and Tools to Research, Plan, and Analyze {#sec-analysis}

With the harness configured, select the capabilities needed for the archaeological task. Use them in a sequence that produces inspectable outputs: source notes, an approved plan, reviewed code, and a rendered report. The researcher decides when the work is ready to advance.

## Select and Test the Capabilities You Need

A skill is a package of instructions and supporting resources for a recurring task [@Xu2026; @Anthropic2025Skills]. A tool is a function the agent can call, such as looking up a DOI or running R. MCP is a standard for connecting an agent to tools and data sources; an MCP server supplies those connections [@Anthropic2024; @Hou2025]. A skill describes a procedure, while tools perform its actions. @tbl-stack lists the five MCP servers and five skills named in this guide.

```{r}
#| label: tbl-stack
#| tbl-cap: "MCP servers and skills installed in the author's OpenCode configuration."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-stack.csv")))
```

Start with a task such as retrieving a cited paper or checking a calculation. Inspect the skill's instructions and the server's documentation before enabling either. Give each tool only the access its task requires, and test its output against a known input. Untrusted documents or tool descriptions can contain instructions that redirect an agent, a risk examined in MCP security research [@Hou2025]. Review unexpected requests for file access, code execution, or network connections before allowing them.

An archaeology team could write a skill that specifies its compendium structure, expected checks, and reporting conventions. Record the skill version and any local modifications with the analysis. A documented procedure makes the intended method inspectable; it does not guarantee that every agent run follows it.

## Research the Question and Check the Sources

Use literature tools to retrieve evidence relevant to the question. Services such as Crossref and Semantic Scholar provide bibliographic records; PDF-conversion tools such as markitdown make documents available as text. Check extracted text against the original when page order, tables, or mathematical notation matter. Literature retrieval and document conversion are separate operations from evaluating the evidence. @Faniel2013 found that reusers judge collection context, wording, and repository trust decisive, and @Huggett2018 warns that preservation alone does not produce reuse, so we check fitness before repurposing any archive.

For each proposed citation, ask the agent to retrieve the record and locate the passage supporting the claim. Check the title, authors, year, and DOI against the publication, then inspect the relevant passage yourself. A real DOI confirms a record's identity, not its relevance to a sentence. @Leung2023's warning about fictitious citations supports checking references regardless of the interface used to generate them. Flag unresolved citations instead of filling gaps with plausible details. Save source notes with the retrieved record, supporting passage, page or section location, and verification status.

The same principle applies when AI extracts variables from archaeological reports. @Hariri2025 describes structured knowledge extraction from archaeological texts, and @Syme2026 examines validation of LLM-generated units for cross-cultural research. @Gilardi2023 found {LLM} labels matched or exceeded crowd workers on tested tasks, while @Ziems2024 found uneven validity across social concepts, so we validate each coding task on a human-coded sample. One LLM extracted ecological data from literature 50 times faster than a human reviewer, with accuracy above 90% on discrete and categorical data [@Gougherty2024]. It performed poorly on certain quantitative data, so quantitative fields need extra quality checks [@Gougherty2024]. GPT-4 extracted species at F1 99.8% and locations at 95.3% from pest-control abstracts and distinguished ecological roles, though a small number of confabulated details appeared [@Scheepens2024]. Context-rich prompting improved every tested model, though no single model dominated all extraction tasks, so test several model-prompt pairings and keep validation workflows that build trust in outputs [@Moorthy2026]. Before using machine-coded variables, define the categories, compare a sample against human coding, and document errors. Save source locations with extracted values so another researcher can inspect the evidence.

## Draft and Critique the Analysis Plan

Write the initial plan yourself. Specify the archaeological question, observation unit, variables, exclusions, treatment of missing data, and proposed methods. Identify the assumptions that would make the analysis appropriate. @Spillias2025 recommends documenting expert knowledge that AI training materials may omit. For an archaeological analysis, record relevant excavation practices, changes in recording conventions, and known gaps in recovery. State each check explicitly: in one study no tested model checked predictor collinearity because the specification never asked, though the check is standard practice [@Brown2026a]. Save the plan as a text file, such as `analysis-plan.md`, so its revisions can be reviewed with Git. Ideation and feedback are uses of generative AI identified by @Korinek2023; we recommend recording which suggestions the researcher accepts or rejects. @Huggett2022 distinguishes augmentation, automatization, and heteromation, so we assign assistive tasks to the software and keep method choice with the researcher.

Ask the agent to identify gaps before generating code: "What information is missing? Which assumptions require checking? What alternative analysis would answer this question?" Evaluate the response against the data documentation and literature. Prompt specificity pays: a generic prompt never elicited count models for abundance data, while a detailed prompt specifying variables, sample size, and design guaranteed them across models [@Brown2026b]. Detailed prompts also produced more consistent code across repeated runs [@Brown2026b]. @Messeri2024 cautions that authoritative explanations can create an illusion of understanding. Require reasons for each proposed method, and set aside suggestions you cannot evaluate.

For example, suppose a plan compares artifact densities between excavation trenches. Before choosing a test, identify whether the records describe artifacts, excavation units, or trench totals. Check how excavation volume was measured and whether units share contexts. Ask the agent to distinguish counts from calculated densities and explain how the proposed method addresses unequal sampling effort. Many zeros alone are insufficient grounds for selecting a quasi-Poisson model. Resolve these questions with statistical guidance appropriate to the design before approving code generation.

A second model can provide another critique, but agreement between models is not independent validation. @Bertran2026 reports that analytical conclusions can vary with model and prompt choices. Record substantive changes to the plan and their reasons. Following @Piper2026's theory-first approach, require each choice to address the archaeological question rather than merely produce an available output. Applicability errors dominate model mistakes: even GPT-4o peaks near 65% on a statistical-analysis benchmark, while humans err by confusing tasks, so model and researcher supply complementary checks [@Zhu2024].

## Execute and Verify the Analysis

Ask the agent to implement one approved part of the plan at a time in the Quarto document, since smaller tasks are easier to learn from and test [@Mammides2024]. Require it to report the changed files, commands run, checks performed, and unresolved problems. Inspect the Git diff before accepting the work. For the artifact-count task, compare plotted totals with independently checked records and confirm that missing labels receive the planned treatment. Test numerical results rather than accepting successful execution as evidence of correctness [@Wilson2014]. Confirm that scripts actually ran and reports actually rendered rather than trusting a completion message: in one controlled study, agents reported finishing tasks they had not completed and filled results tables with fabricated values after code silently failed [@Brown2026a]. Inspect the diff for scope as well as correctness: agents sometimes add analyses or files nobody requested, which look thorough but were not asked for [@Brown2026book].

Tools that execute R can expose errors for inspection, but automated correction still needs review. @Brown2026a tested agentic workflows for ecological and fisheries modelling and found variable accuracy and failures to follow instructions. When code fails, check whether the proposed correction preserves the approved method. Do not accept a different analysis simply because it runs. Agents also change methods without flagging it: one agent swapped a requested bootstrap interval for a t-distribution interval, so read the code produced, not the summary [@Brown2026a]. Such limits also distinguish this workflow from self-improving research pipelines: The AI Scientist generated, debugged, and selected among its own experiments through automated multi-stage search, with humans filtering outputs rather than reviewing each code change [@Lu2026]. We recommend the reverse arrangement, in which the agent reports changes and waits for approval.

Check whether plausible analytical alternatives change the conclusion. Multiverse analysis evaluates results across a defined set of defensible choices [@Steegen2016]. Recent agent studies emphasize how model and prompt choices can affect analytical exploration [@Miao2026; @Bertran2026]. For an archaeological analysis, specify alternatives with substantive justification, document the variants actually tested, and report disagreements. An agent can help execute these checks, but the researcher must define the defensible choices and budget their computational cost. Before trusting a model on real data, fit the same code to simulated data with a chosen effect [@Brown2026book]. If the pipeline cannot recover a relationship built in on purpose, distrust it on data where the answer is unknown. Repeat the simulation and check that the true value falls inside the 95% interval in most runs. Check row counts before and after every join or filter: an unexpected change usually means a many-to-many join mistaken for a one-to-one match [@Brown2026book]. Sanity-check summaries against known ranges: a maximum far outside expectation signals a units error, not a finding. As an option, write the check before the code: assert a coefficient sign known from prior work in a testthat test, watch it fail, then implement until it passes [@Brown2026book].

## Render and Archive the Results

Render the Quarto document from a fresh R session and inspect the text, figures, tables, and references together. Record package versions with renv, then test the documented procedure in a separate checkout of the project. A successful render verifies execution in that environment; interpretation still requires the checks described above.

Prepare a research compendium containing the permitted data, source document, analysis plan, project instructions, dependency record, and reviewed code [@Marwick2017]. Standardized quality-control guidelines for archived data and code now exist for ecology and evolution, motivated by persistently low reproducibility of archived materials [@Pick2026]. Record the model identifier, dates, relevant prompts, and tool or skill versions where available. Reporting standards for LLM-assisted work now exist: reviewers use the recorded methods to judge rigor and replicability [@Abdurahman2025]. Remove credentials and restricted material before sharing any logs. Provide a README with the render command, expected outputs, data provenance, licenses, and access restrictions. Archive an approved release in a repository such as Zenodo or the Open Science Framework and cite its DOI. @Wilkinson2016 defines {FAIR} stewardship, @Carroll2020 adds {CARE} governance, and @Gupta2023 applies both to {Canadian} archaeology, so we record consent and access terms with the deposit. @Wells2026 emphasizes that archaeological archiving also requires attention to data labor and Indigenous data sovereignty.

The output of this practice is an inspected analysis and the materials needed to reproduce its saved results. @fig-workflow summarizes how preparation and harness configuration support that work.

```{r}
#| label: fig-workflow
#| echo: false
#| fig-cap: "Three practices for agentic AI-assisted archaeological analysis. Researcher review connects preparation, harness configuration, and the research-to-archive sequence."
#| fig-alt: "Three boxes run from left to right: learn R, Quarto, and Git; configure a harness with instructions and permissions; use skills and tools to research, plan, execute, verify, and archive. A return arrow indicates revision after checks. Researcher approval applies throughout."
#| fig-width: 8
#| fig-height: 4

par(mar = c(1, 1, 1, 1))
plot.new()
plot.window(xlim = c(0, 12), ylim = c(0, 5))

nodes <- list(
  list(x = 1.8, label = "1. Learn tools\nand standards\nR, Quarto, Git", fill = "#D9EAF5"),
  list(x = 5.7, label = "2. Configure\nthe harness\nInstructions, permissions", fill = "#FCE8BE"),
  list(x = 9.8, label = "3. Use skills\nand tools\nResearch, plan, execute\nVerify, archive", fill = "#D9EFE7")
)

draw_node <- function(node) {
  rect(node$x - 1.65, 1.6, node$x + 1.65, 3.7,
       col = node$fill, border = "grey30", lwd = 1.2)
  text(node$x, 2.65, labels = node$label, cex = 0.85)
}

invisible(lapply(nodes, draw_node))
arrows(c(3.5, 7.4), c(2.65, 2.65), c(4, 8.1), c(2.65, 2.65),
       length = 0.08, lwd = 1.5)
arrows(9.8, 1, 5.7, 1, length = 0.08, lty = 2, col = "grey30")
text(7.75, 0.55, "Revise after checks", cex = 0.8)
text(6, 4.5, "Researcher approval throughout", cex = 1, font = 2)
```

# A Checklist Before You Publish

Use @tbl-checklist to review the completed workflow. Record unresolved checks before treating the report as ready for circulation.

```{r}
#| label: tbl-checklist
#| tbl-cap: "Checks before publishing an AI-assisted archaeological analysis."
#| echo: false
knitr::kable(read.csv(here::here("analysis", "figures", "table-checklist.csv")))
```

# Discussion

The three practices assign distinct responsibilities to the researcher. Familiarity with R, Quarto, and Git supports inspection of calculations, reports, and changes. Harness configuration defines the permitted working environment. Tested skills and tools then support research, planning, execution, and verification within that environment. This sequence applies existing scientific-computing practices to agent-assisted archaeological work [@Wilson2014; @Marwick2017; @Bridgeford2026].

Each practice has costs. Learning R and version control requires time, while configuring a harness introduces provider choices and access decisions. Skills and servers add dependencies that require testing and maintenance. @Kansa2026 discusses commercial dependence, and @JohnsonEberl2026 examines environmental costs of AI in archaeology. @Luccioni2024 measured inference energy across models and tasks, so we batch queries and log usage. Treat heavy token use as a reason to inspect output: in one study the costliest runs were agents stuck in unproductive loops, not the most accurate ones [@Brown2026a]. Mapping latency, token cost, and F1 score guides model choice under budget constraints [@Moorthy2026]. We recommend recording available cost and usage information, limiting unnecessary repeated calls, and considering whether an existing R script already completes the task.

These instructions do not establish that agent assistance improves every analysis. The illustrative tasks are teaching examples, and we have not measured their effects on accuracy or completion time. Reproducing saved R code differs from repeating an agent conversation: a model may generate different code on another attempt. Provider-side model updates add a second source of non-repeatability: the same prompt can behave differently months later because the provider changed the model [@Abdurahman2025; @Abramson2026]. Preserve the inspected code and document the interaction, rather than promising identical agent behavior. Incentives sharpen a further risk: when publication metrics define success, researchers may use LLMs as a shortcut to game output rather than support innovation, which reduces quality and stifles development [@Johnson2024].

# Conclusion

We recommend three practices for responsible agentic AI use in archaeological data analysis: learn the research tools and standards, configure a harness with explicit permissions, and use tested skills and tools to research, plan, and analyze. Quarto makes the relationship between code and report inspectable, while Git records reviewed changes. Planning and verification remain part of the researcher's work throughout.

The practical result should be a documented analysis whose inputs, decisions, code, and outputs another researcher can examine. The agent may assist with each task, but the researcher must explain the methods, assess the evidence, and authorize release. These responsibilities define when the work is ready to share.

# Acknowledgments

No permits were required for the preparation of this manuscript. This article describes general practices for using AI tools in archaeological data analysis. The author used OpenCode (https://opencode.ai/), an open-source AI coding agent, and OpenRouter (https://openrouter.ai/) to assist with drafting, code verification, and reference checking during the preparation of this manuscript. All AI-generated suggestions were reviewed and revised by the author.

# Funding Statement

This research received no specific grant funding from any funding agency, commercial or not-for-profit sectors.

# Data Availability Statement

No original data were used.

# Competing Interests

The author declares none.

# Notes

This manuscript was prepared as a Quarto document within an rrtools research compendium. The rendered output files (paper.docx) and the Colophon section at the end of this document are internal to the compendium and should be removed before submission to a journal.

# References

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
